Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2012-07-30 MIMOSA.tex
Web supp changes.. added derivations and references
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+? Didn't look at CD8's}. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see Web Figure A).

@@

All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}). An R package will soon be released as part of the Bioconductor project\footnote{Add reference}.

@@

%	&=\int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u\\

@@

%	\intertext{The integrand is the kernel of a beta distribution with parameters $\left(n_u+n_s+\alpha_u,N_u+N_s+\beta_u\right)$, giving a closed form expression for the  posterior predictive distribution of model \eqref{eq:null}:}
%	\mathrm{Pr}(y_i|\alpha_u,\beta_u)&=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1postsupp}\\
%	\intertext{with marginal log-likelihood:}
%	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N^i_{s}+n^i_{s}}{n^i_{s}}}+\log{\binom{N^i_{u}+n^i_{u}}{n^i_{u}}}+\right.\\
%	\left.\log{\left(\mathrm{B}(n^i_s+n^i_u+\alpha_u,N^i_s+N^i_u+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLLsupp}

@@

\bibliographystyle{mn2e}

>>>>

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see Web Figure A).

@@

All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}). An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.

@@

%&=\int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u\\
%\begin{split}

@@

%\end{split}\\
%\intertext{The integrand is the kernel of a beta distribution with parameters $\left(n_u+n_s+\alpha_u,N_u+N_s+\beta_u\right)$, giving a closed form expression for the  posterior predictive distribution of model \eqref{eq:null}:}
%\mathrm{Pr}(y_i|\alpha_u,\beta_u)&=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1postsupp}\\
%\intertext{with marginal log-likelihood:}
%\begin{split}
%\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N^i_{s}+n^i_{s}}{n^i_{s}}}+\log{\binom{N^i_{u}+n^i_{u}}{n^i_{u}}}+\right.\\
%\left.\log{\left(\mathrm{B}(n^i_s+n^i_u+\alpha_u,N^i_s+N^i_u+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLLsupp}

@@

\bibliographystyle{biom}
